---
title: "Data_processing_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data_processing_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette uses a publically available dataset containing human kidney biopsy RNAseq from DKD patients (GSE142025) to perform an example data analysis in R using supervisoR and DESeq2. 

```{r setup}
library(supervisoR)
library(DESeq2)
library(dplyr)
library(magrittr)
library(readr)
library(tibble)
library(piano)
library(tidyr)
```

# Loading data

We load data directly from GEO. This code has been modified from the GEO2R website (https://www.ncbi.nlm.nih.gov/geo/geo2r) to simplify the data processing. 

```{r}
# load counts table from GEO
urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
path <- paste(urld, "acc=GSE142025", "file=GSE142025_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep="&");
tbl <- as.matrix(read_tsv(path)) %>% `rownames<-`(., .[,1]) %>% .[,-1]

# load gene annotations 
apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
annot <- read_tsv(apath, quote="")
rownames(annot) <- annot$GeneID

# sample selection
gsms <- "000000000000000000000111111222222222"
sml <- strsplit(gsms, split="")[[1]]

# group membership for samples
gs <- factor(sml)
groups <- make.names(c("aDKD","eDKD","Healthy"))
levels(gs) <- groups
sample_info <- data.frame(Group = gs, row.names = colnames(tbl))

# pre-filter low count genes
# keep genes with at least N counts > 10, where N = size of smallest group
keep <- rowSums( tbl >= 10 ) >= min(table(gs))
tbl <- tbl[keep, ]

ds <- DESeqDataSetFromMatrix(countData=tbl, colData=sample_info, design= ~Group)

ds <- DESeq(ds, test="LRT", reduced = ~ 1)  # Use LRT for all-around gene ranking

# extract results
r_aDKD <- results(ds, contrast = c("Group", "aDKD", "Healthy"), alpha=0.05, pAdjustMethod ="BH") %>% as.data.frame()
r_eDKD <- results(ds, contrast = c("Group", "eDKD", "Healthy"), alpha=0.05, pAdjustMethod ="BH") %>% as.data.frame()

r_all <- left_join(r_aDKD %>%
                     rownames_to_column("gene") %>%
                     dplyr::select(gene, log2FoldChange),
                   r_eDKD %>%
                     rownames_to_column("gene") %>%
                     dplyr::select(gene, log2FoldChange),
                   by = "gene",
                   suffix = c(".aDKD", ".eDKD")) %>%
  left_join(.,
            annot %>%
              as.data.frame() %>%
              dplyr::select(GeneID, Symbol) %>%
              mutate_all(as.character),
            by = c("gene" = "GeneID")) %>%
  dplyr::select(!gene) %>%
  dplyr::group_by(Symbol) %>%
  dplyr::summarize(across(where(is.numeric), ~ mean(.x ))) %>%
  ungroup() %>%
  column_to_rownames("Symbol") %>%
  `colnames<-`(., sub("log2FoldChange.","",colnames(.)))
```

# Run pathway analysis
We use the piano package with fgsea to run the pathway analysis. Here we are using the included MSigDB version of the Reactome pathways, but this could be replaced with any pathway database. It should however have some hierarchical element for proper use with the supervisoR package. Take a look at the introduction vignette to see how you could use your own database for supervisoR visualization

```{r}
## We load the pathway database directly from the supervisoR package!
default_data <- supervisoR::load_default_data(database = "reactome",
                                  species = "Homo sapiens")

res_list <- NULL
myGSC <- piano::loadGSC(stack(default_data$gene_sets))
for (i in colnames(r_all)) {
  tmp_lfc <- r_all[i] %>%
    mutate_all(~ replace_na(.x, 1))
  
  piano_res <- piano::runGSA(geneLevelStats = tmp_lfc, 
    gsc = myGSC, geneSetStat = "fgsea", signifMethod = "geneSampling", 
    nPerm = 10000, adjMethod="BH", gsSizeLim = c(5, Inf))
  
  tmp_res <- piano::GSAsummaryTable(piano_res) %>%
      dplyr::select(Name,
                    `Stat (dist.dir)`,
                    `p adj (dist.dir.up)`,
                    `p adj (dist.dir.dn)`) %>%
      `colnames<-`(c("pathway",
                     "statDistinctDir",
                     "pAdjDistinctDirUp",
                     "pAdjDistinctDirDn")) %>%
      mutate(contrast = i) %>%
      mutate_at(vars(pAdjDistinctDirUp,pAdjDistinctDirDn), ~ if_else(is.na(.), 1, .)) %>%
      mutate(direction = case_when(pAdjDistinctDirUp > pAdjDistinctDirDn ~ "up",
                                   pAdjDistinctDirUp < pAdjDistinctDirDn ~ "down",
                                   T ~ "neither")) 
  
    res_list[[i]] <- tmp_res
}

pw_out <- do.call(rbind, unname(res_list)) %>%
  dplyr::select(pathway, statDistinctDir, contrast) %>%
  pivot_wider(id_cols = pathway,
              names_from = contrast,
              values_from = statDistinctDir) %>%
  column_to_rownames("pathway")
```

# Visualize using supervisoR
We can now visualize our results using both the plotting functionality and the shiny app. 
```{r}
p1 <- plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  mapping = default_data$mapping,
  g = default_data$pathway_graph,
  gene_sets = default_data$gene_sets,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME,
  enrichment_scores = pw_out,
  conditions = colnames(pw_out),
  layout = "dh",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = F,
  max_depth = 2
)

run_pathway_shiny_app(mapping = default_data$mapping,
  g = default_data$pathway_graph,
  gene_sets = default_data$gene_sets,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME,
  enrichment_scores = pw_out,
  conditions = colnames(pw_out),
  enrichment_limits = NULL,
  glyph_size = c(80, 60),
  res = 96,
  hide_nodes_without_enrichment = TRUE,
  layout_options_plot = list(
    layout = "layout_with_kk",
    physics = FALSE,
    smooth = FALSE,
    type = "full",
    randomSeed = 42,
    options = NULL
  )
)
```
