---
title: "Introduction to pathway visualization with supervisoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pathway visualization with supervisoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  fig.retina = 3, 
  out.width = "100%"
)
```

# Introduction

supervisoR is an R package designed for visualizing the enrichment of biological pathways and/or genes across conditions. This vignette explains the basic functionalities of the package, guiding you through data preparation and various visualization techniques.

# Installing and loading the package

```{r setup}
library(supervisoR)
library(magrittr)
```

# Data preparation
## Loading Example Enrichment Scores

```{r}
# First we define which database and species we want to look at. Currently, the "reactome" and "GO" are supported through the MsigDB databases. Human ("Homo sapiens") and ("Mus musculus") are currently supported. However, custom data can be provided - see the final part of this vignette for an example using the included data. We recommend that you use one of the provided gene_sets for your pathway analysis for easy interoperability. 

database = "reactome"
species = "Homo sapiens"

# Load the default included data to generate example dataset
default_data <- load_default_data(database = database, species = species)


# Generate example dataset. Note that the example data is randomly generated using rnorm(length(gene_sets), sd = 0.25) and thus does not reflect real conditions or output from a pathway analysis.
enrichment_scores <- data.frame(Condition.1 = rnorm(length(default_data$gene_sets), sd = 0.25),
                                Condition.2 = rnorm(length(default_data$gene_sets), sd = 0.25),
                                Condition.3 = rnorm(length(default_data$gene_sets), sd = 0.25),
                                Condition.4 = rnorm(length(default_data$gene_sets), sd = 0.25))
rownames(enrichment_scores) <- names(default_data$gene_sets)
conditions <- colnames(enrichment_scores)
```

# Visualizing Pathways

## Plotting a Subgraph with Glyphs

```{r}
# Define the parent pathway name. Select an example below depending on whether it is a reactome or GO graph.
parent_name <- "Signaling By Tgfb Family Members" # REACTOME
# parent_name <- "Polysaccharide Metabolic Process" #GO

# Plotting multiple conditions
p1 <- plot_subgraph(
  parent_geneset_name = parent_name,
  database = database,
  species = species,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE
)

print(p1)
```

## Plotting multiple conditions relative to Condition 1

You can visualize enrichment scores relative to a specific reference condition by specifying the reference parameter.

```{r}
# Plotting multiple conditions relative to Condition 1
p2 <- plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  database = database,
  species = species,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1"
)

print(p2)
```

## Adjusting Edge Thickness and Adding Percentage Labels

Enhance the visualization by adjusting edge thickness based on gene overlap and adding percentage labels to edges.

```{r}
# Plotting multiple conditions relative to Condition 1 with edge thickness and labels
p3 <- plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  database = database,
  species = species,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1",
  adjust_edge_thickness = TRUE,
  edge_percentage_labels = TRUE
)

print(p3)
```

## Plotting a Subgraph with a Single Condition

Focus on a single condition to explore specific pathway enrichments without the complexity of multiple comparisons.

```{r}
# Plotting a subgraph for a single condition
p4 <- plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  enrichment_scores = enrichment_scores,
  conditions = "Condition.3",
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  database = database,
  species = species,
  layout = "fr",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = TRUE
)

print(p4)
```
## Plotting a Subgraph with a Single Condition, up to three links from the root

Focus on a single condition to explore specific pathway enrichments without the complexity of multiple comparisons. Here we limit the graph depth to 2

```{r}
# Plotting a subgraph for a single condition. Limit the depth of the graph to 2 to limit the confusion from too many nodes.
p5 <- plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  enrichment_scores = enrichment_scores,
  conditions = "Condition.3",
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  database = database,
  species = species,
  layout = "fr",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = TRUE,
  max_depth = 2
)

print(p5)
```


## Using a shiny app to explore the full pathway set.

If you don't know where to start, a shiny app is available to explore the pathway set interactively. This includes most of the options found in the static plots.

```{r, out.width = "200%"}
run_pathway_shiny_app(enrichment_scores,
  conditions,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME
  database = database,
  species = species,
  enrichment_limits = NULL,
  glyph_size = c(80, 60),
  res = 96,
  layout_options_plot = list(
    layout = "layout_with_kk",
    physics = FALSE,
    smooth = FALSE,
    type = "full",
    randomSeed = 42,
    options = NULL
  )
)
```

# Using custom pathway data
## Creating custom pathway graph, gene_sets and mapping. Here, the default package data is used but any data of a similar format should work
```{r}
# Load the default data for analyses (MsigDB REACTOME dataset or MsigDB GOBP terms). This data is already preprocessed, but we will use it as an example of how the data should look.
default_data <- load_default_data(database = "reactome", species = "Homo sapiens")

# Extract components from the loaded data
gene_sets <- default_data$gene_sets
relations <- default_data$relations

# Create translation layer mapping 
trans <- data.frame(
  gene_sets_name = default_data$mapping$processed_name,
  relation_name = default_data$mapping$exact_source,
  stringsAsFactors = FALSE
)

# Preprocess gene sets using the mapping and relations 
# In order to process the data we require two things; a list of genesets and a dataframe of pathway relations. The geneset list must contain named lists of genes, with the names corresponding to the geneset. The pathway relations dataframe must contain a column named "parent" and "child", indicating which nodes in the hierarchy should be linked and how. An optional third relation column describes the relation - e.g. for GO pathways you can have a pathway regulate another.
# A translation_layer dataframe is required if the pathway names in you relation dataframe does not match those in the geneset dataframe. It must contain two columns, gene_sets_name and relation_name.
preprocessed_data <- load_and_preprocess_gene_sets(
  gene_sets = gene_sets,
  pathways_relation = relations,
  translation_layer = trans
)

# Unpack the preprocessed data
gene_sets <- preprocessed_data$gene_sets
mapping <- preprocessed_data$mapping
g <- preprocessed_data$pathway_graph

p6 <- plot_subgraph(
  parent_geneset_name = parent_name,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  pathway_column = NULL, # WE SET THIS TO NULL SINCE WE HAVE THE PATHWAYS AS ROWNAMES. OTHERWISE, SPECIFY THE COLUMN BY EITHER INDEX OR NAME,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE
)

print(p6)

```

# Conclusion

This vignette demonstrated how to use supervisoR's functions to visualize pathway enrichment data. The unified plot_subgraph function provides a flexible and efficient way to generate visualizations for a specific subset of pathways. the run_pathway_shiny_app function provides a broader, interactive exploration of pathway regulation.
