---
title: "Introduction to pathway visualization with supervisoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pathway visualization with supervisoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

supervisoR is an R package designed for visualizing the enrichment of biological pathways and/or genes across conditions. This vignette explains the basic functionalities of the package, guiding you through data preparation and various visualization techniques.

# Installing and loading the package

```{r setup}
library(supervisoR)
library(magrittr)
```

# Data preparation

## Loading Gene Sets and building Pathway Graph

```{r}
# Load the default data for analyses (MsigDB REACTOME dataset or MsigDB GOBP terms). This data is already preprocessed.
default_data <- load_default_data(database = "reactome", species = "Homo sapiens")

# Extract components from the loaded data
gene_sets <- default_data$gene_sets
relations <- default_data$relations

# Create translation layer mapping (EXAMPLE, NOT NECESSARY WHEN USING DEFAULT DATA)
trans <- data.frame(
  gene_sets_name = default_data$mapping$processed_name,
  relation_name = default_data$mapping$exact_source,
  stringsAsFactors = FALSE
)

# Preprocess gene sets using the mapping and relations (EXAMPLE, NOT NECESSARY WHEN USING DEFAULT DATA)
preprocessed_data <- load_and_preprocess_gene_sets(
  gene_sets = gene_sets,
  pathways_relation = relations,
  translation_layer = trans
)

# Unpack the preprocessed data
gene_sets <- preprocessed_data$gene_sets
mapping <- preprocessed_data$mapping
g <- preprocessed_data$pathway_graph
```

## Loading Example Enrichment Scores

```{r}
# Generate example dataset. Note that the example data is randomly generated using rnorm(length(gene_sets), sd = 0.25) and thus does not reflect real conditions or output from a pathway analysis.
enrichment_scores <- data.frame(Condition.1 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.2 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.3 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.4 = rnorm(length(gene_sets), sd = 0.25))
rownames(enrichment_scores) <- names(gene_sets)
conditions <- colnames(enrichment_scores)
```

# Visualizing Pathways

## Plotting a Subgraph with Glyphs

```{r}
# Define the parent pathway name. Select an example below depending on whether it is a reactome or GO graph.
parent_name <- "Signaling By Tgfb Family Members" # REACTOME
#parent_name <- "Polysaccharide Metabolic Process" #GO
# Plotting multiple conditions
plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE
)
```

## Plotting multiple conditions relative to Condition 1

You can visualize enrichment scores relative to a specific reference condition by specifying the reference parameter.

```{r}
# Plotting multiple conditions relative to Condition 1
plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1"
)
```

## Adjusting Edge Thickness and Adding Percentage Labels

Enhance the visualization by adjusting edge thickness based on gene overlap and adding percentage labels to edges.

```{r}
# Plotting multiple conditions relative to Condition 1 with edge thickness and labels
plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1",
  adjust_edge_thickness = TRUE,
  edge_percentage_labels = TRUE
)
```

## Plotting a Subgraph with a Single Condition

Focus on a single condition to explore specific pathway enrichments without the complexity of multiple comparisons.

```{r}
# Plotting a subgraph for a single condition
plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  enrichment_scores = enrichment_scores,
  conditions = "Condition.3",
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "fr",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = TRUE
)
```

## Using a shiny app to explore the full pathway set.

If you don't know where to start, a shiny app is available to explore the pathway set interactively. This includes most of the options found in the static plots.

```{r}
run_pathway_shiny_app(enrichment_scores,
  conditions,
  mapping,
  g,
  gene_sets,
  enrichment_limits = NULL,
  glyph_size = c(80, 60),
  res = 96,
  layout_options_plot = list(
    layout = "layout_with_kk",
    physics = FALSE,
    smooth = FALSE,
    type = "full",
    randomSeed = 42,
    options = NULL
  ) 
)
```

# Conclusion

This vignette demonstrated how to use supervisoR's functions to visualize pathway enrichment data. The unified plot_subgraph function provides a flexible and efficient way to generate visualizations for a specific subset of pathways. the run_pathway_shiny_app function provides a broader, interactive exploration of pathway regulation.
