---
title: "Introduction to pathway visualization with supervisoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pathway visualization with supervisoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 10,
  fig.retina = 3, 
  out.width = "100%"
)
```

# Introduction

supervisoR is an R package designed for visualizing the enrichment of biological pathways and/or genes across conditions. This vignette explains the basic functionalities of the package, guiding you through data preparation and various visualization techniques.

# Installing and loading the package

```{r setup}
library(supervisoR)
library(magrittr)
```

# Data preparation

## Loading Gene Sets and building Pathway Graph

```{r}
# Load the default data for analyses (MsigDB REACTOME dataset or MsigDB GOBP terms). This data is already preprocessed.
default_data <- load_default_data(database = "reactome", species = "Homo sapiens")

# Extract components from the loaded data
gene_sets <- default_data$gene_sets
relations <- default_data$relations

# Create translation layer mapping (EXAMPLE, NOT NECESSARY WHEN USING DEFAULT DATA)
trans <- data.frame(
  gene_sets_name = default_data$mapping$processed_name,
  relation_name = default_data$mapping$exact_source,
  stringsAsFactors = FALSE
)

# Preprocess gene sets using the mapping and relations (EXAMPLE, NOT NECESSARY WHEN USING DEFAULT DATA)
# In order to process the data we require two things; a list of genesets and a dataframe of pathway relations. The geneset list must contain named lists of genes, with the names corresponding to the geneset. The pathway relations dataframe must contain a column named "parent" and "child", indicating which nodes in the hierarchy should be linked and how. An optional third relation column describes the relation - e.g. for GO pathways you can have a pathway regulate another.
# A translation_layer dataframe is required if the pathway names in you relation dataframe does not match those in the geneset dataframe. It must contain two columns, gene_sets_name and relation_name.
preprocessed_data <- load_and_preprocess_gene_sets(
  gene_sets = gene_sets,
  pathways_relation = relations,
  translation_layer = trans
)

# Unpack the preprocessed data
gene_sets <- preprocessed_data$gene_sets
mapping <- preprocessed_data$mapping
g <- preprocessed_data$pathway_graph
```

## Loading Example Enrichment Scores

```{r}
# Generate example dataset. Note that the example data is randomly generated using rnorm(length(gene_sets), sd = 0.25) and thus does not reflect real conditions or output from a pathway analysis.
enrichment_scores <- data.frame(Condition.1 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.2 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.3 = rnorm(length(gene_sets), sd = 0.25),
                                Condition.4 = rnorm(length(gene_sets), sd = 0.25))
rownames(enrichment_scores) <- names(gene_sets)
conditions <- colnames(enrichment_scores)
```

# Visualizing Pathways

## Plotting a Subgraph with Glyphs

```{r}
# Define the parent pathway name. Select an example below depending on whether it is a reactome or GO graph.
parent_name <- "Signaling By Tgfb Family Members" # REACTOME
# parent_name <- "Polysaccharide Metabolic Process" #GO

# Plotting multiple conditions
p1 <- plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE
)

print(p1)
```

## Plotting multiple conditions relative to Condition 1

You can visualize enrichment scores relative to a specific reference condition by specifying the reference parameter.

```{r}
# Plotting multiple conditions relative to Condition 1
p2 <- plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1"
)

print(p2)
```

## Adjusting Edge Thickness and Adding Percentage Labels

Enhance the visualization by adjusting edge thickness based on gene overlap and adding percentage labels to edges.

```{r}
# Plotting multiple conditions relative to Condition 1 with edge thickness and labels
p3 <- plot_subgraph(
  parent_geneset_name = parent_name,
  enrichment_scores = enrichment_scores,
  conditions = conditions,
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "kk",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  reference = "Condition.1",
  adjust_edge_thickness = TRUE,
  edge_percentage_labels = TRUE
)

print(p3)
```

## Plotting a Subgraph with a Single Condition

Focus on a single condition to explore specific pathway enrichments without the complexity of multiple comparisons.

```{r}
# Plotting a subgraph for a single condition
p4 <- plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  enrichment_scores = enrichment_scores,
  conditions = "Condition.3",
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "fr",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = TRUE
)

print(p4)
```
## Plotting a Subgraph with a Single Condition, up to three links from the root

Focus on a single condition to explore specific pathway enrichments without the complexity of multiple comparisons. Here we limit the graph depth to 2

```{r}
# Plotting a subgraph for a single condition
p5 <- plot_subgraph(
  parent_geneset_name = "Metabolism Of Lipids",
  enrichment_scores = enrichment_scores,
  conditions = "Condition.3",
  mapping = mapping,
  g = g,
  gene_sets = gene_sets,
  layout = "fr",
  hide_nodes_without_enrichment = TRUE,
  enrichment_limits = NULL,
  use_node_label = TRUE,
  repel = TRUE,
  max_depth = 2
)

print(p5)
```


## Using a shiny app to explore the full pathway set.

If you don't know where to start, a shiny app is available to explore the pathway set interactively. This includes most of the options found in the static plots.

```{r, out.width = "200%"}
run_pathway_shiny_app(enrichment_scores,
  conditions,
  mapping,
  g,
  gene_sets,
  enrichment_limits = NULL,
  glyph_size = c(80, 60),
  res = 96,
  layout_options_plot = list(
    layout = "layout_with_kk",
    physics = FALSE,
    smooth = FALSE,
    type = "full",
    randomSeed = 42,
    options = NULL
  )
)
```

# Conclusion

This vignette demonstrated how to use supervisoR's functions to visualize pathway enrichment data. The unified plot_subgraph function provides a flexible and efficient way to generate visualizations for a specific subset of pathways. the run_pathway_shiny_app function provides a broader, interactive exploration of pathway regulation.
